The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Tides or nature?

### 2025-08-31 thesis/introduction.tex
thesis julio ch1-3 edits

Dwarf galaxies host, in many ways, the most extreme galactic
environments in the universe. These galaxies are typically defined to be
fainter than the Large Magellanic Cloud (LMC), with \(M_V \gtrsim -18\)
or similarly \(M_\star \lesssim 10^9 M_\odot\) {[}e.g.,
\citet{hodge1971}; \citet{mcconnachie2012}). Because the galaxy
luminosity function increases towards fainter objects, dwarfs are the
most numerous galaxies in the Universe. Dwarf galaxies are also highly
\emph{dark-matter dominated}, with mass to light ratios which may exceed
1000 \(M_\odot/ L_\odot\). With the exception of the Magellanic Cloud,
all dwarf galaxy satellites of the Milky Way (MW) are \emph{quenched},
with little to no recent star formation. Indeed, most faint MW
satellites contain stellar populations which are \emph{relics} from the
early universe, consisting of many of the oldest and most metal poor
stars.

Understanding the properties of dwarf galaxies thus has implications
across astronomy, from cosmological structure formation on small scales
to the formation of metal-poor stellar populations.

\emph{Remove this paragraph?} Of the classical dwarfs, we find that the
Sculptor and Ursa Minor dwarf spheroidal galaxies stand out with a more
extended density profile relative to an exponential, hinting at tidal
effects. As a case study in tidal effects and our understanding of dwarf
galaxy formation and evolution, we aim to understand the dynamical
history of these galaxies.

In this section, we first describe the general observed properties of
local dwarf galaxies. Next, we summarize the cosmological origin and
role of dwarf galaxies. We then review recent updates to the observed
properties of dwarf galaxies made on new facilities. We then discuss how
tides influence satellite galaxies in general. Finally, we provide a
roadmap to the thesis. \textbf{Look at this paragraph again}

@@

The earliest spectroscopic work hinted that dwarf galaxies may contain
substantial amounts of dark matter. From early determinations of the
velocity dispersion for the Sculptor and Ursa Minor dwarf spheroidal
(dSph) galaxies \citep[e.g.,][]{aaronson1983, aaronson+olszewski1987},
inferred mass-to-light ratios were at least 10 times larger than the
values of globular clusters (GCs). While rather uncertain initially,
these values have been corroborated with larger and more precise samples
\citep[e.g.,][]{hargreaves+1994}. At the time, several theories were
proposed to explain these unusually high mass-to-light ratios. Examples
include: ongoing tidal disruption artificially inflating the
line-of-sight velocity dispersions \citep[e.g.,][]{kuhn+miller1989}, the
presence of massive central black holes
\citep[e.g.,][]{strobel+lake1994}, or modified theories of gravity
\citep{milgrom1995}. Over time, however, a consensus developed where the
high mass-to-light ratios of dwarf galaxies was due to the presence of a
dark matter halo, which was lacking in GCs
\citep[e.g.,][\textbf{REFS}]{dekel+silk1986}. Since then, the properties
of dwarf galaxies have played an increasingly important role in our
understanding of the clustering of dark matter on small scales {[}e.g.,
sales+2022, bullock+boylan-kolchyn2017{]}.

Dwarf galaxies span a large range of physical sizes, luminosities, and
morphologies. Broadly, there are three classes of dwarf galaxies based
on luminosity, as illustrated by Fig.~\ref{fig:galaxy_images}. Local
\textbf{bright dwarfs} with magnitudes
\(-18 \lesssim M_V \lesssim  -14\), often exhibit irregular morphologies
and recent star formation. Fig.~\ref{fig:galaxy_images} shows the Large
Magellanic Cloud (LMC) as an example of an irregular, bright dwarf
galaxy, where most stars are in a rotationally-supported thin disk (seen
nearly face-on) with a prominent bar. \textbf{Classical dwarf
spheroidals} occupy intermediate luminosities (
\(-14 \lesssim M_V  \lesssim -7.7\)). Typically, these systems are old,
non-star forming, gas-poor, and spheroidal. All Milky Way satellites
discovered before digital sky surveys are classicals. The 12 classical
dwarf satellites of our Galaxy are (in order of decreasing luminosity)
Sagittarius, Fornax, Leo I, Sculptor, Antlia II, Leo II, Carina, Draco,
Ursa Minor, Canes Venatici I, Sextans I, and Crater II.\footnote{While
  formally the dwarf galaxy names we discuss contain ``dwarf
  spheroidal'' (dSph), e.g.~Sculptor dSph, we omit this suffix for
  brevity.} The \textbf{ultra-faint}s dwarfs (UFDs) occupy the very
faintest magnitude range (\(-7.7 \lesssim M_V\)). These galaxies have
minuscule stellar masses, tend to be more compact, and include the
faintest known galaxies (\textbf{ref}). Altogether, known dwarf galaxies
span more than 15 orders in magnitude, or over 7 decades in stellar
mass. \textbf{Rewrite to include stellar mass ranges as well}

A common definition today for a dwarf galaxy is a stellar system with
evidence for dark matter (high mass-to-light ratios, e.g.~\textbf{REF}).
In contrast, star clusters (like GCs) have no clear evidence for dark
matter.

\begin{figure}
\centering
\includegraphics[width=5.41667in,height=5.41667in]{figures/galaxy_pictures.png}
\caption[Dwarf Galaxy Pictures]{Images of the LMC (DSS2), Fornax (DES
DR2), Sculptor (DES DR2), and Ursa Minor (UNWISE with Gaia point sources
over-plotted). Each image includes the galaxy's luminosity and a 200 pc
scale bar. \emph{TODO: Mark half-light radius with
circle?}}\label{fig:galaxy_images}
\end{figure}

@@

We only understand a tiny fraction of the universe's composition. The
leading theory of cosmology, \LCDM (Lambda Cold Dark Matter), posits
that the universe is composed of about 68\% dark energy (\(\Lambda\)),
27\% dark matter (DM), and 5\% regular baryons\footnote{Astronomers like
  to change definitions of words. \emph{Baryons} here means
  baryons+leptons, i.e.~any standard model massive fermion.}
\citep{planckcollaboration+2020}. While the composition of dark matter
and dark energy remains elusive, we know their general properties. Dark
energy causes the acceleration of the expansion of the universe on large
scales. We do not discuss dark energy here---it does not substantially
affect the Local Group today. Dark matter, instead, makes up the vast
majority of mass in galaxies. Typically, galaxies have baryonic to dark
matter ratios of between 1:5 to beyond 1:1000 for faint dwarf galaxies.
In \LCDM, dark matter is assumed to interact only gravitationally. Light
passes through dark matter unimpeded---in this sense dark matter is
transparent. Dark matter is also commonly assumed to be \emph{cold},
i.e.~typical velocities much smaller than the speed of light in the
early universe. If dark matter is cold, then it should condense on all
scales, forming non-linear structures (or \emph{halos}) from the size of
galaxy clusters to smaller than the faintest dwarf galaxies.
Implications of dark matter properties range from cosmological
structural formation, galaxy structure, and galaxy interactions.

\textbf{is this ordered logically?}

\subsection{\texorpdfstring{Dwarf galaxy formation in
\LCDM}{Dwarf galaxy formation in }}\label{dwarf-galaxy-formation-in}

\textbf{In the next section, we consider galaxy formatin in LCDM}

@@

Dark matter halos, and their associated galaxies, rarely evolve in
isolation. Instead, structure formation is \emph{hierarchical}. Small
dark matter halos collapse first and hierarchically merge into
progressively larger halos
\citep[e.g.,][]{blumenthal+1984, white+rees1978, white+frenk1991}.
Structure formation happens on a wide range of scales. The largest
structures become the cosmic web---composed of voids, filaments, and
sheets---and the smallest non-linear structures (halos, \textbf{define
non-linear}) host dwarf galaxies. Hierarchical assembly is evident
through the large scale structure of the universe, remnants of past
mergers within the Milky Way, and tidal disruption of dwarf galaxies and
their streams around nearby galaxies.

Small-scale structure formation is sensitive to deviations from
\LCDM cosmology \citep[e.g.,][]{bechtol+2022}. One key prediction of
\LCDM is that mass perturbations are expected to exist on all scales,
and are largest on the smallest scales, so we would expect the formation
of halos on all scales. Many alternative models, such as warm dark
matter, may smooth out small-scale features and reduce the abundance of
small halos or change their structure \citep[e.g.,][]{lovell+2014}.
Dwarf galaxies, which occupy the smallest dark matter halos, are
promising windows into small-scale structures. To understand the
evolution of these objects, we need to understand the general
predictions of \LCDM for the properties of dwarf galaxies.

@@

In \LCDM cosmological simulations, dark matter halos are remarkably
self-similar. In \citet{NFW1996, NFW1997}, hereafter NFW, the authors
observe that the spherically-averaged density profiles \(\rho(r)\) are
universally well described by a two-parameter law:
\begin{equation}\protect\phantomsection\label{eq:nfw}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
}\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
density . This profile has shown remarkable success in describing
\LCDM halos across several orders of magnitude in mass. NFW profiles are
\emph{cuspy}, where the density continuously rises like
\(\rho \sim 1/r\) at small radii \(r \ll r_s\). The steepness of the
density profile increases gradually with radius, and at large radii the
density falls off like \(\rho \sim 1/r^3\). Fig.~\ref{fig:nfw_density}
shows an example NFW halo.

The total mass of an NFW profile formally diverges, so halo masses are
commonly defined using an overdensity criterion. The virial mass,
\(M_{200}\), is defined as the mass within a radius, \(r_{200}\),
containing a mean enclosed density 200 times\footnote{For the collapse
  of a uniform spherical density, the virialized overdensity would be
  \(\Delta = 18\pi^2\approx 178\) for a critical universe
  \(\Omega_m = 1\). This is commonly rounded to \(\Delta = 200\). While
  this parameter may be closer to \(\Delta \approx 100\) for our
  universe, \(\Delta\) also increases with redshift \citep[using eq. 6
  from][]{bryan+norman1998}.} the critical density of the universe:
\begin{equation}{
M_{200} =200\,\frac{4\pi}{3} \ r_{200}^3\ \rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z)^2 / 8\pi G
}\end{equation} A standard second parameter of NFW halos is the halo
concentration, \(c=r_{200} / r_s\) which describes how the
characteristic radius scale of the halo compares to the virial radius.
Using this parameter, the scale density is a function of \(c\) alone,
\(\rho_s = (200/3)\,\rho_{\rm crit} c^3 / [\log(1+c) - c/(1+c)]\)
\citep{NFW1996}.

An equivalent, alternative characterization of NFW halos uses their
circular velocity profiles. The circular velocity,
\(\V_{\rm circ}(r) = \sqrt{G M(r) / r}\), reaches a maximum
\(\V_{\rm max}\) at radius \(r_{\rm max} \approx 2.16258\,r_s\).
\(\V_{\rm max}\) and \(r_{\rm max}\), like \(M_{200}\) and \(c\), fully
specify an NFW halo,

The two parameters of an NFW profile are not independent. Lower mass
dark matter halos often collapse earlier, when the universe was denser.
As a result, low mass subhalos tend to be more concentrated
\citep[e.g.,][]{NFW1997}. The relationship between \(M_{200}\) and c, or
the mass-concentration relation, describes the mean trend of
concentration with mass \citep[e.g.,][]{bullock+2001, ludlow+2016}. The
left panel of Fig.~\ref{fig:smhm} illustrates the present-day
mass-concentration relation in terms of \(r_{\rm max}\) and
\(\V_{\rm max}\). While there is a general trend for concentration to
increase with mass, the concentration values still have substantial
scatter. Other parameters such as the halo spin or shape may affect the
scatter of the mass-concentration relation, but their effect is
typically expected to be small
\citep{navarro+2010, dicintio+2013, dutton+maccio2014}.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/example_density_profiles.pdf}}
\caption[Example density profiles]{Density profiles in log density
versus log radius for stars and dark matter in a Sculptor-like galaxy.
The dark matter is more extended and massive than the star across the
entire galaxy.}\label{fig:nfw_density}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/cosmological_means.pdf}}
\caption[Stellar-mass halo-mass relation]{\textbf{Left} Radius of
maximum circular velocity \(r_{\rm max}\) as a function of maximum
circular velocity. The solid line with 1-\(\sigma\) shaded region and
dashed line are the relations from \citet{ludlow+2016} for \(z=0\) and
\(z=2\) respectively. \textbf{Right} Stellar mass (top) as a function of
maximum circular velocity. The solid line with the 1-\(\sigma\) shaded
region is the relation from \citet{fattahi+2018} with scatter points
simulated central galaxies from APOSTLE in
\citet{fattahi+2018}.}\label{fig:smhm}
\end{figure}

\subsection{\texorpdfstring{Formation of stars within
\LCDM halos}{Formation of stars within halos}}\label{formation-of-stars-within-halos}

The observed abundance of galaxies may be compared with the abundance of
\LCDM halos to derive constraints regarding which galaxies inhabit which
halos. This technique, dubbed ``abundance matching,'' implies a tight
relation between the stellar mass of a galaxy and the mass of the halo
it inhabits \citep{li+white, moster+naab+white2013}.

Fig.~\ref{fig:smhm} shows the stellar mass versus halo mass relation
(SMHM, with halo mass represented by \(\V_{\rm max}\)) predicted by the
\LCDM cosmological hydrodynamical simulations from the the APOSTLE
project \citep{sawala+2016}, which uses the hydrodynamical setup from
the EAGLE project \citep{crain+2015, schaye+2015} to simulate Local
Group analogues in a \LCDM cosmological context \textbf{(db: reorder
these sentences)}. While there is some scatter, the range of predicted
\({\rm v}_{\rm max}\) is fairly narrow across \(\sim 4\) decades in
stellar mass. This figure indicates that the SMHM relation becomes
increasingly steep in the dwarf galaxy regime---many dwarf galaxies are
formed in relatively massive halos. Because lower mass galaxies have
increasingly shallow potential wells, feedback becomes more effective at
removing gas. Re-ionization additionally suppresses late star formation
in the faintest galaxies. As a result, the resulting stellar mass of a
galaxy is highly sensitive to the dark matter mass, especially for faint
dwarf galaxies. If we know the initial stellar mass of a dwarf galaxy,
we also know, at some level, the properties of its dark matter halo.
\textbf{fix this paragraph}

In \LCDM galaxy formation, the majority of mass in dwarf galaxies comes
from Dark Matter. Given the SMHM, Fig.~\ref{fig:nfw_density} shows an
example stellar component for a classical dwarf galaxy with the expected
underlying dark matter halo. Even when the stars are most dense in the
inner regions, the dark matter remains nearly an order of magnitude
higher in density. Stars have a small contribution to the gravitational
structure of dwarf galaxies--instead stars are reasonably approximated
as tracer particles of the underlying dark matter halo. \textbf{add
something about stellar compactness}

Several challenges complicate the SMHM trend, including environment,
assembly history, tidal effects, and the details of galaxy formation. In
particular, effects like ram-pressure stripping (removal of gas in the
dwarf galaxy due to pressure from host's circumgalactic medium) and
tidal removal of gas cause star formation to quench {[}\textbf{REFS}{]}.
Additionally, the time of formation (relative to reionization) can
influence the resulting stellar content \citep{kim+2024}. Finally, tides
influence both the dark matter and stellar mass but in different amounts
\citep[e.g.,][]{PNM2008}. Consequently, tides may reduce the halo mass
more than the stellar mass, adding additional scatter to the SMHM trend,
particularly for low-mass satellites \citep[e.g.,][]{fattahi+2018}.
Understanding the effects of tides on Local Group dwarf galaxies is
essential to determine where and how these galaxies formed in a
cosmological context.

@@

Since Local Group dwarfs are nearby, they are resolved into individual
stars, and we can study these galaxies on a star-by-star basis. As a
result, it is possible to measure the 3D velocity and position of a star
if we can measure the its position, distance, line-of-sight (LOS)
velocity, and proper motions. Unfortunately, determining distances and
full 3D velocities is challenging. The most direct measurement of
distance, parallax, requires precise tracking of a star's sky position
across a year. And while line-of-sight (LOS) velocities are relatively
easily determined from spectroscopy, tangential velocities, derived from
proper motions and distances, are much more challenging. Typically,
measuring proper motions requires accurate (much less than arcsecond)
determinations of small changes in a star's position over baselines of
several years to decades. The full 6D position and velocity information
for stars has, until recently, been known for only a handful of stars.

@@

The Local Group (LG) is defined as galaxies which are within \(\sim 1\)
Mpc from the Milky Way-Andromeda system \citep[e.g.,][ and references
therein]{mcconnachie2012}. \emph{Gaia} has revolutionized LG studies.
For example, the 6D dynamical measurements and metallicities of MW stars
led to the (re)discovery of past mergers or Milky Way building blocks
like \emph{Gaia}-Sausage Enceladus
\citetext{\citealp[e.g.,][]{helmi+2018}; \citealp{belokurov+2018}; \citealp[but
see also][]{mesa+2005}}, out-of-equilibrium structures like the
\emph{Gaia} snail \citep[e.g.,][]{antoja+2018}, and dynamical effects of
the Milky Way's spiral arms and the bar in the solar neighbourhood
\citep{hunt+vasiliev2025}. Moving to the Milky Way halo, \emph{Gaia} has
helped find and constrain numerous stellar streams
\citep[\citet{bonaca+price-whelan2025}]{ibata+malhan+martin2019}.
Altogether, \emph{Gaia} has revealed the hierarchical formation and
complex, evolving structure of our own Galaxy.

>>>>

Dwarf galaxies host, in many ways, the most extreme galactic
environments in the universe. These galaxies are typically defined to be
fainter than the Large Magellanic Cloud (LMC), with \(M_V \gtrsim -18\)
or similarly \(M_\star \lesssim 10^9 M_\odot\)
\citep[e.g.,][]{hodge1971, mcconnachie2012}. Because the galaxy
luminosity function increases towards fainter objects, dwarfs are the
most numerous galaxies in the Universe
\citep[e.g.,][]{blanton+2005, mao+2021}. Dwarf galaxies are also highly
\emph{dark-matter dominated}, with mass to light ratios which may exceed
1000 \(M_\odot/ L_\odot\) \citep[e.g.,][]{simon+geha2007, hayashi+2023}.
With the exception of the Magellanic Clouds, most dwarf galaxy
satellites of the Milky Way (MW) are \emph{quenched}, with little to no
recent star formation \citep[e.g.,][]{weisz+2014}. Indeed, most faint MW
satellites contain stellar populations which are \emph{relics} from the
early universe, consisting of many of the oldest and most metal poor
stars. Understanding the properties of dwarf galaxies thus has
implications across astronomy, from cosmological structure formation on
small scales to the formation of metal-poor stellar populations.

In this section, we first describe the general observed properties of
local dwarf galaxies. Next, we summarize the cosmological origin and
role of dwarf galaxies. We later review recent advancements and problems
concerning dwarf galaxies. Next, we contextualize the extended stellar
density profiles of Sculptor and Ursa Minor. We then discuss how tides
influence satellite galaxies in general. Finally, we provide a roadmap
to the thesis.

@@

The earliest spectroscopic work hinted that dwarf galaxies may contain
substantial amounts of dark matter. From early determinations of the
velocity dispersion for the Sculptor and Ursa Minor dwarf spheroidal
(dSph) galaxies \citep[e.g.,][]{aaronson1983, aaronson+olszewski1987},
inferred mass-to-light ratios were at least 10 times larger than the
values of globular clusters (GCs). While rather uncertain initially,
these values have been corroborated with larger and more precise samples
\citep[e.g.,][]{hargreaves+1994}. At the time, several theories were
proposed to explain these unusually high mass-to-light ratios. Examples
include: ongoing tidal disruption inflating inferred velocity
dispersions \citep[e.g.,][]{kuhn+miller1989}, the presence of massive
central black holes \citep[e.g.,][]{strobel+lake1994}, or modified
theories of gravity \citep{milgrom1995}. Over time, however, a consensus
developed where the high mass-to-light ratios of dwarf galaxies was due
to the presence of a dark matter halo, which was lacking in GCs
\citep[e.g.,][]{dekel+silk1986, wechsler+tinker2018}. Since then, the
properties of dwarf galaxies have played an increasingly important role
in our understanding of the clustering of dark matter on small scales
\citep[e.g.,][]{sales+2022, bullock+boylan-kolchin2017}.

Today, a common definition for a dwarf galaxy is a gravitationally bound
stellar system with dark matter \citep[or, more specifically, not
consistent with newtonian dynamics of visible matter
alone,][]{willman+strader2012}. In contrast, star clusters (like GCs)
have no clear evidence for dark matter. The boundary between these two
classes becomes uncertain for faint, compact systems, known as
``ambiguous'' systems.

The Local Group (LG) is defined as the group consisting of galaxies
within \(\sim 1\) Mpc from the Milky Way-Andromeda centre \citep[e.g.,][
and references therein]{mcconnachie2012}. Today, we know that the Milky
Way system is teeming with satellites.
Fig.~\ref{fig:mw_satellite_system} shows the MW satellite system,
including dwarf galaxies, globular clusters, and ambiguous systems. Such
a nearby population of dwarf galaxies are useful for resolved, detailed
studies aiming to understand the nature of these systems.

Dwarf galaxies span a large range of physical sizes, luminosities, and
morphologies. Broadly, there are three classes of dwarf galaxies based
on luminosity, as illustrated by Fig.~\ref{fig:galaxy_images}. Local
\textbf{bright dwarfs} with magnitudes
\(-18 \lesssim M_V \lesssim  -14\) or stellar masses
\(10^9\,\Mo \gtrsim M_\star \gtrsim 3\times10^7\,\Mo\) \footnote{assuming
  stellar mass-to-light ratio of 1, may be \(\sim 2\) for older
  populations}, often exhibit irregular morphologies and recent star
formation. Fig.~\ref{fig:galaxy_images} shows the Large Magellanic Cloud
(LMC) as an example of an irregular, bright dwarf galaxy, where most
stars are in a rotationally-supported thin disk (seen nearly face-on)
with a prominent bar. \textbf{Classical dwarf spheroidals} occupy
intermediate luminosities ( \(-14 \lesssim M_V  \lesssim -7.7\) or
\(3\times10^7\,\Mo \gtrsim M_\star \gtrsim 10^5\,\Mo\)). Typically,
these systems are old, non-star forming, gas-poor, and spheroidal. All
Milky Way satellites discovered before digital sky surveys are
classicals. The 12 classical dwarf satellites of our Galaxy are (in
order of decreasing luminosity) Sagittarius, Fornax, Leo I, Sculptor,
Antlia II, Leo II, Carina, Draco, Ursa Minor, Canes Venatici I, Sextans
I, and Crater II.\footnote{While formally the dwarf galaxy names we
  discuss contain ``dwarf spheroidal'' (dSph), e.g.~Sculptor dSph, we
  omit this suffix for brevity.} The \textbf{ultra-faint}s dwarfs (UFDs)
occupy the very faintest magnitude range (\(-7.7 \lesssim M_V\), or
\(M_\star < 10^5\,\Mo\)). These galaxies have minuscule stellar masses,
typically compact sizes, and very metal-poor stellar populations
\citep[see review][]{simon2019}. Altogether, known dwarf galaxies span
more than 15 orders in magnitude, or over 7 decades in stellar mass.

Here, we focus on classical dwarf galaxies, which are ideal for detailed
dynamical study. These galaxies are dark matter dominated but also have
large numbers of stars with radial velocities and proper motions, and
have generally well-constrained properties.

\begin{figure}
\centering
\includegraphics[width=5.41667in,height=5.41667in]{figures/galaxy_pictures.png}
\caption[Dwarf Galaxy Pictures]{Images of the LMC (DSS2), Fornax
\citep[DES DR2,][]{abbott+2021}, Sculptor (DES DR2), and Ursa Minor
\citep[UNWISE,][with \textit{Gaia} point sources
over-plotted]{lang2014, meisner+lang+schlegel2017, meisner+lang+schlegel2017a}.
The grey ellipse represents the half-light radius for the three dwarf
spheroidals, and the luminosity is derived from the absolute V-band
magnitude of each galaxy.}\label{fig:galaxy_images}
\end{figure}

@@

We only understand a fraction of the universe's composition. The leading
theory of cosmology, \LCDM{} (Lambda Cold Dark Matter), posits that the
universe is composed of about 68\% dark energy (\(\Lambda\)), 27\% dark
matter (DM), and 5\% regular baryons\footnote{\emph{Baryons} here means
  baryons+leptons, i.e.~any standard model massive fermion.}
\citep{planckcollaboration+2020}. While the composition of dark matter
and dark energy remains elusive, we know their general properties. Dark
energy drives the acceleration of the expansion of the universe on large
scales. We do not discuss dark energy here---it does not substantially
affect the Local Group today. Dark matter, instead, makes up the vast
majority of mass in galaxies. Typically, galaxies have baryonic to dark
matter ratios of between 1:5 to beyond 1:1000 for faint dwarf galaxies
\citep[e.g.,][]{hayashi+2023}.

In \LCDM{}, dark matter is assumed to interact only gravitationally.
Light passes through dark matter unimpeded---in this sense dark matter
is transparent. Dark matter is also commonly assumed to be \emph{cold},
i.e.~typical velocities much smaller than the speed of light in the
early universe. If dark matter is cold, then it should condense on all
scales, forming non-linear structures (or \emph{halos}) from the size of
galaxy clusters to smaller than the faintest dwarf galaxies.
Implications of dark matter properties include cosmological structural
formation, galaxy structure, and galaxy interactions.

\subsection{\texorpdfstring{Structure formation in
\LCDM{}}{Structure formation in }}\label{structure-formation-in}

@@

Dark matter halos, and their associated galaxies, rarely evolve in
isolation. Instead, structure formation is \emph{hierarchical}. Small
dark matter halos collapse first and hierarchically merge into
progressively larger halos
\citep[e.g.,][]{blumenthal+1984, white+rees1978, white+frenk1991}.
Hierarchical assembly is evident through the large scale structure of
the universe, remnants of past mergers within the Milky Way, and tidal
disruption of dwarf galaxies and their streams around nearby galaxies.

Small-scale structure formation is sensitive to deviations from \LCDM{}
cosmology \citep[e.g.,][]{bechtol+2022}. One key prediction of \LCDM{}
is that mass perturbations are expected to exist on all scales, and are
largest on the smallest scales, so we would expect the formation of
halos on all scales. Many alternative models, such as warm dark matter,
may smooth out small-scale features and reduce the abundance of small
halos or change their structure \citep[e.g.,][]{lovell+2014}. Dwarf
galaxies, which occupy the smallest dark matter halos, are promising
windows into the behaviour of dark matter on small scales.

@@

In \LCDM{} cosmological simulations, dark matter halos are remarkably
self-similar. In \citet{NFW1996, NFW1997}, hereafter NFW, the authors
observe that the spherically-averaged density profiles \(\rho(r)\) are
universally well described by a two-parameter law,
\begin{equation}\protect\phantomsection\label{eq:nfw}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
}\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
density. This profile has shown remarkable success in describing \LCDM{}
halos across several orders of magnitude in mass. NFW profiles are
\emph{cuspy}, where the density rises like \(\rho \sim 1/r\) at small
radii \(r \ll r_s\). The steepness of the density profile increases
gradually with radius, and at large radii the density falls off like
\(\rho \sim 1/r^3\). Fig.~\ref{fig:nfw_density} shows an example NFW
halo.

The total mass of an NFW profile formally diverges, so halo masses are
commonly defined using an overdensity criterion. The virial mass,
\(M_{200}\), is defined as the mass within a radius, \(r_{200}\),
containing a mean enclosed density 200 times\footnote{For the collapse
  of a uniform spherical density, the virialized overdensity would be
  \(\Delta = 18\pi^2\approx 178\) for a critical universe
  \(\Omega_m = 1\). This is commonly rounded to \(\Delta = 200\). While
  this parameter may be closer to \(\Delta \approx 100\) for our
  universe, \(\Delta\) also increases with redshift \citep[using eq. 6
  from][]{bryan+norman1998}.} the critical density of the universe:
\begin{equation}{
M_{200} =200\,\frac{4\pi}{3} \ r_{200}^3\ \rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z)^2 / 8\pi G.
}\end{equation} A standard second parameter of NFW halos is the halo
concentration, \(c=r_{200} / r_s\), which describes how the
characteristic radius scale of the halo compares to the virial radius.
Using this parameter, the scale density is a function of \(c\) alone,
\(\rho_s = (200/3)\,\rho_{\rm crit} c^3 / [\log(1+c) - c/(1+c)]\)
\citep{NFW1996}.

An equivalent, alternative characterization of NFW halos uses their
circular velocity profiles. The circular velocity,
\(\vcirc(r) = \sqrt{G M(r) / r}\), reaches a maximum \(\vmax\) at radius
\(\rmax \approx 2.16258\,r_s\). \(\vmax\) and \(r_{\rm max}\), like
\(M_{200}\) and \(c\), fully specify an NFW halo.

The two parameters of an NFW profile are not independent. Lower-mass
dark matter halos often collapse earlier, when the universe was denser.
As a result, low mass subhalos tend to be more concentrated
\citep[e.g.,][]{NFW1997}. The relationship between \(M_{200}\) and c, or
the mass-concentration relation, describes the mean trend of
concentration with mass \citep[e.g.,][]{bullock+2001, ludlow+2016}. The
left panel of Fig.~\ref{fig:smhm} illustrates the present-day
mass-concentration from \citet{ludlow+2016}. While concentration tends
to decrease with increasing mass, the relation has substantial scatter.
Other parameters such as the halo spin or shape may affect the scatter
of the mass-concentration relation, but their effect is typically
expected to be small
\citep{navarro+2010, dicintio+2013, dutton+maccio2014}.

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/example_density_profiles.png}
\caption[Example density profiles]{Density profiles in log density
versus log radius for stars and dark matter in a Sculptor-like galaxy.
The dark matter is more extended and massive than the star across the
entire galaxy.}\label{fig:nfw_density}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/cosmological_means.pdf}}
\caption[Stellar-mass halo-mass relation]{\textbf{Left} The NFW halo
concentration \(c=r_{200} / r_s\) as a function of virial mass
\(M_{200}\). The solid line with 1-\(\sigma\) shaded region is the
mass-concentration relation from \citet{ludlow+2016} for \(z=0\).
\textbf{Right} Stellar mass (top) as a function of maximum circular
velocity. The solid line with the 1-\(\sigma\) shaded region is the
relation from \citet{fattahi+2018} with scatter points simulated central
galaxies from \apostle{} in \citet{fattahi+2018}.}\label{fig:smhm}
\end{figure}

\subsection{\texorpdfstring{Formation of galaxies within \LCDM{}
halos}{Formation of galaxies within  halos}}\label{formation-of-galaxies-within-halos}

The observed abundance of galaxies may be compared with the abundance of
\LCDM{} halos to derive constraints regarding which galaxies inhabit
which halos. This technique, dubbed ``abundance matching,'' implies a
tight relation between the stellar mass of a galaxy and the mass of the
halo it inhabits \citep{li+white2009, moster+naab+white2013}.

Fig.~\ref{fig:smhm} shows the stellar mass versus halo mass relation
(SMHM, with halo mass represented by \(\vmax\)) predicted by the \LCDM{}
cosmological hydrodynamical simulations of Local Group analogues from
the \apostle{} project \citep{sawala+2016}.\footnote{\apostle{}
  simulated Local Group analogues in a \LCDM{} cosmological context with
  the hydrodynamical setup from the \eagle{} simulations
  \citep{crain+2015, schaye+2015}}. While there is some scatter, the
range of predicted \(\vmax\) is fairly narrow across \(\sim 4\) decades
in stellar mass. This figure indicates that the SMHM relation becomes
increasingly steep in the dwarf galaxy regime---many dwarf galaxies are
formed in relatively massive halos. Because lower mass galaxies have
increasingly shallow potential wells, feedback becomes more effective at
removing gas. Re-ionization additionally suppresses late star formation
in the faintest galaxies. As a result, the resulting stellar mass of a
dwarf galaxy is highly sensitive to the dark matter halo mass.

In \LCDM{} galaxy formation, the majority of mass in a dwarf galaxies
comes from the extended dark matter halo. Fig.~\ref{fig:nfw_density}
shows an example exponential stellar component for a classical dwarf
galaxy with the expected underlying dark matter halo. Even when the
stars are most dense in the inner regions, the dark matter remains
nearly an order of magnitude higher in density. Stars have a small
contribution to the gravitational structure of dwarf galaxies--instead
stars are reasonably approximated as tracer particles of the underlying
dark matter halo. In addition, the stellar component is much more
compact than the dark matter halo.

Several challenges complicate the SMHM trend, including environment,
assembly history, tidal effects, and the details of galaxy formation. In
particular, effects like ram-pressure stripping (removal of gas in the
dwarf galaxy due to pressure from host's circumgalactic medium) and
tidal removal of gas cause star formation to quench
\citep[e.g.,][]{christensen+2024}. Additionally, the time of formation
(relative to reionization) can influence the resulting stellar content
\citep{kim+2024}. Finally, tides influence both the dark matter and
stellar mass but in different amounts \citep[e.g.,][]{PNM2008}.
Consequently, tides may reduce the halo mass more than the stellar mass,
adding additional scatter to the SMHM trend, particularly for low-mass
satellites \citep[e.g.,][]{fattahi+2018}. Understanding the effects of
tides on Local Group dwarf galaxies is essential for determining where
and how these galaxies formed in a cosmological context.

@@

Since Local Group dwarfs are nearby, they are resolved into individual
stars, and we can study these galaxies on a star-by-star basis. As a
result, it is possible to measure the 3D velocity and position of a star
if we can measure the its position, distance, line-of-sight (LOS)
velocity, and proper motions. Unfortunately, determining distances and
full 3D velocities is challenging. The most direct measurement of
distance, parallax, requires precise tracking of a star's sky position
across a year. And while line-of-sight (LOS) velocities are relatively
easily determined from spectroscopy, tangential velocities, derived from
proper motions and distances, are much more challenging. Typically,
measuring proper motions requires accurate (much less than arcsecond)
determinations of small changes in a star's position over baselines of
years to decades. The full 6D position and velocity information for
stars has, until recently, been known for only a handful of stars.

@@

\emph{Gaia} has revolutionized LG studies. For example, the 6D dynamical
measurements and metallicities of MW stars led to the (re)discovery of
past mergers or Milky Way building blocks like \emph{Gaia}-Sausage
Enceladus
\citetext{\citealp[e.g.,][]{helmi+2018}; \citealp{belokurov+2018}; \citealp[but
see also][]{meza+2005}}, out-of-equilibrium structures like the
\emph{Gaia} snail \citep[e.g.,][]{antoja+2018}, and dynamical effects of
the Milky Way's spiral arms and the bar in the solar neighbourhood
\citep[ and references therein]{hunt+vasiliev2025}. Moving to the Milky
Way halo, \emph{Gaia} has helped find and constrain numerous stellar
streams \citep{ibata+malhan+martin2019, bonaca+price-whelan2025}.
Altogether, \emph{Gaia} has revealed the hierarchical formation and
complex, evolving structure of our own Galaxy.
